---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PairwiseChiSquared

<!-- badges: start -->
<!-- badges: end -->

The goal of **PairwiseChiSquared** is to facilitate the analysis of categorical data across multiple samples using the chi-squared test. It first enables you to perform an overall chi-squared test to determine if *any* differences are present, then to perform a set of pairwise chi-squared tests to determine *exactly which* differences are present.

## Installation

You can install the development version of PairwiseChiSquared from [GitHub](https://github.com/) with:

Note: if it is not already installed, remove the comment sign (#) at the start of the first line to install `devtools`.

``` {r installation}
# install.packages("devtools", dependencies = TRUE)
devtools::install_github("RobCrawford527/PairwiseChiSquared", dependencies = TRUE)
```

The `dependencies = TRUE` argument ensures that other packages that are required for proteomicshelpers to work correctly are installed too.

## Data preparation

To use the chi-squared test your data must be categorical - i.e. they are counts of values in a number of non-overlapping, mutually exclusive categories. For example, this might be numbers of granules per cell from microscopy images, binned into counts of 0, 1-2, 3-4 and 5+ granules per cell. The selection of bin boundaries is important: the general rule is that *≥80% of your categories should have expected values ≥5*. If this is not the case and many categories have low or zero values, consider changing the boundaries to combine categories.

To use the PairwiseChiSquared functions, your data must be in the format of a table, with your **samples as columns** and your **bins as rows**. The value in each cell should be the **count** of measurements in each bin for the appropriate sample. It is fine if the total numbers of measurements for the samples differ: if this is the case you should *leave the counts as is*, do not convert them to percentages.

You can prepare the table in e.g. Excel, export it as a `.txt` file and then import to R. Make sure the bins are set as the *row names* rather than a separate column. Once imported, the data frame should look similar to the example below:

```{r example}
example_data <- data.frame(Sample1 = c(10,20,30),
                           Sample2 = c(10,12,38),
                           Sample3 = c(10,22,28),
                           row.names = c("A", "B", "C"))
example_data
```




You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
